(working title) Local Physiological Adaptation in \textit{Mimulus cardinalis}

### 2015-04-21 ms/ms.tex
New git ignore and vastly updated ms folder

\documentclass[11pt, oneside]{article}

@@

@@

@@

%
% Start document
%

\begin{document}
\maketitle
\listoffigures
\listoftables

@@

@@

% What's the big picture (this is TOO big):
% It is taken for granted that genes are the foundation of evolution. If we get to the genetic basis of adaptive evolution, then we have understood the process at its deepest level. I believe this view is misguided. Genetic change is, to the best of our knowledge, the basis of evolution, but this does not imply that by understanding genetic variation we can predict phenotypic evolution. Rather, variation and constraint on phenotypic evolution are often determined by what goes on outside. Organisms must survive and reproduce in the real world of physical laws and interactions with other organisms. Physiology, the study of organismal function, tells us about the variation and constraint on ways in which organisms can meet these challenges. ...

Local adaptation is one of the most ubiquitous observations in nature: organisms perform well in their natal environment, but poorly outside it. Local adaptation is most often examined within species where selection in different environments leads to ecologically relevant trait divergence between populations. Despite local adaptation leading to population divergence, many traits are conserved within species. \cdm{this is not quite right. What I really want to say is that some traits divergence genetically, others are plastic, but plasticity is conserved because...? why? no spatial variation in selection for plasticity?}. Can we predict which types of traits are likely to be different between populations and which are likely to be constant? A long-standing hypothesis is that the grain of environmental variation determines whether organisms respond through population genetic differentiation or phenotypic plasticity \citep{}. \cdm{What I want to get to is whether plasticity is conserved. The problem is, just because there is selection for plasticity, this doesn't need to imply that the reaction norm is actually the same across populations. But perhaps we can oversimplify here?}.

When the environment is fine grained

, Despite this common, and long-standing pattern, only certain traits In particular, adaptation to the local abiotic environment is generally thought to be the major cause of natural selection. Physiology, the study of organismal function, connects fitness to the abiotic environment, yet we actually know little about the physiological basis of adaptation to different environments within a species. To address this question, we looked at physiological variation across a broad latitudual gradient within \textit{Mimulus cardinalis}...

1. `intrinsic' physiological variation
2. variation in stress response
3. what ecological factors explain divergence (i.e. putative selective agents underlying local adaptation)

 % one paragraph general review of physiology, environment, and local adaptation

 % one paragraph review of physiology, environment, and local adaptation in Mimulus

 % one paragraph wrapping up, motivating this study

@@

I chose 16 populations from throughout the range of \textit{M. cardinalis} (Table 1) that have been previously studied. Seeds were collected in the field \cdm{Amy, can you explain seed collection methods?}.

\begin{table}[htbp]
   \centering
   \topcaption{Table 1: 16 Focal populations}
   \begin{tabular}{@{} lllllll @{}}
      \toprule
      Name    & Region & Demo? & Pop gen? & Lat & Lon & Alt \\
      \midrule
	Hauser Creek						& South Margin		& yes	& yes	& 32.65822	& -116.53235	& 892 \\
	Cottonwood Creek					& South Margin		& yes	& no		& 32.80122	& -116.50194	& 1206 \\
	Sweetwater River					& South Margin		& yes	& yes	& 32.89928	& -116.5849	& 1223 \\
	Grade Road - to Palomar Mountain SP	& South Margin		& no		& no		& 33.31392	& -116.87129	& 1500 \\
	Whitewater Canyon					& Transverse		& yes	& yes	& 33.99329	& -116.66267	& 696 \\
	Mill Creek							& Transverse		& yes	& no		& 34.07808	& -116.87558	& 1992 \\
	West Fork Mojave River				& Transverse		& yes	& no		& 34.28425	& -117.37539	& 1087 \\
	North Fork Middle Tule				& South Sierras		& yes	& yes?	& 36.20081	& -118.65092	& 1284 \\
	Paradise Creek						& South Sierras		& yes	& yes	& 36.51776	& -118.75877	& 887 \\
	Redwood Creek					& South Sierras		& yes	& yes	& 36.69096	& -118.90961	& 1683.72 \\
	Wawona							& Central Sierras	& yes	& yes	& 37.539		& -119.654	& 1208 \\
	Rainbow Pool (RP)					& Central Sierras	& yes	& no		& 37.8188		& -120.00743	& 862.2792 \\
	Middle Yuba (Oregon Creek)			& North Sierras		& yes	& yes	& 39.39442	& -121.08302	& 425.196 \\
	Little Jameson						& North Sierras		& yes	& yes	& 39.74298	& -120.70401	& 1592 \\
	Deep Creek						& North Coast		& yes	& yes	& 41.66546	& -123.11341	& 694 \\
	Rock Creek (North Fork Umpqua)		& North Margin		& yes	& yes	& 43.37375	& -122.9575	& 311.2008 \\
	\bottomrule
   \end{tabular}
\end{table}

@@

% Spring 2014 cohort
On 14 April, 2014, 3-5 seeds per family were sown directly on sand (Quikrete Play Sand, Georgia, USA) watered to field capacity in RLC4 Ray Leach cone-tainers placed in RL98 98-well trays (Stuewe \& Sons, Inc., Oregon, USA). We used pure sand both to facilitate root-washing and because \textit{M. cardinalis} typically grows in sandy, riparian soils (A. Angert, pers. obs.). Two jumbo-sized cotton balls at the bottom of cone-tainers prevented sand from washing out. Cone-tainers were continuously bottom-watered during germination by placing them in medium-sized flow trays (FLOWTMD, Stuewe \& Sons, Inc., Oregon, USA) filled part way with water, placed on benches in greenhouses at the University British Columbia campus in Vancouver, Canada (49\degree 15' N, 123\degree 15' W). Misters thoroughly wetted the top of the sand every two hours during the day. Most seeds germinated between 1 and 2 weeks, but we allowed 3 weeks before transferring seedlings to growth chambers. Germination was recorded daily from one to two weeks after sowing, and every few days thereafter. On 5 May (21 days after sowing), seedlings were transferred to one of two MODEL Growth Chambers (Conviron, Manitoba, Canada). We thinned seedlings to one plant per cone-tainer, leaving the center-most, largest plant. 702 of 768 (91.4\%) had plants that could be used in the experiment. We allowed one week at constant, non stressful conditions (day: 20\celsius, night: 16\celsius) for plants to acclimate to growth chambers before starting treatments. The initial size of seedlings, measured as the length of the first true leaves, did not differ between populations, families, or treatments (Table S\#).

@@

To do this, I will use 4 watering treatments on two populations on two soils. The watering treatments are: constant water, constant level/decreasing frequency, decreasing level/constant frequency, no water. I will monitor soil water potential in pots with no plants. Basically, I'm just trying to figure out whether one or another method is easier or more effective, and if I see any obvious response from plants in term of growth.

@@

We simulated typical growing season (June and July) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (High Temp) and Little Jameson (Low Temp). We downloaded daily interpolated mean, minimum, and maximum air temperature from 13 years (2000-2012) at both sites from ClimateWNA (CITE). Daily temperatures from ClimateWNA are usually highly correlated with the air temperature recorded from data loggers in the field at these sites (A. Angert, unpub. data). Hence, the ClimateWNA temperature profiles are likely to similar to actual thermal regimes experienced by \textit{M. cardinalis} in nature. To create realistic temperature regimes, we calculated the mean temperature trend from June to July using LOESS \citep{Cleveland_etal_1992}. The residuals were highly autocorrelated at both sites (warmer than average days are typically followed by more warm days) and there was strong correlation ($r = 0.65$) between sites (warm days in WWC were also warm in LIJ). The `VARselect' function in the vars package for R \citep{Pfaff_2008} indicated that a lag two Vector Autoregression (VAR(2)) model best captured the within-site autocorrelation as well as between-site correlation in residuals. We fit and simulated from the VAR(2) using the package dse \citep{Gilbert_2006?} in R. Simulated data closely resembled the autocorrelation and between-site correlation of the actual data. From simulated mean temperature, we next selected minimum and maximum daily temperatures. Mean, min, and max temperature were highly correlated at both sites. We chose min and max temperatures using site-specific fitted linear models between mean, max, and min temperature, with additional variation given by normally-distributed random deviates with variance equal to the residual variance of the linear models. For each day, the nighttime (22:00 - 6:00) chamber temperature was set to the simulated minimum temperature. During the middle of the day, chamber temperature was set to the simulated maximum temperature, with a variable period of transition between min and max so that the average temperature was equal the simulated mean temperature.

@@

			Are there any data from California about soil water content through the season?

\subsection*{Monitoring environment}

\subsection*{Traits}

\subsubsection*{Structure}

leaf traits, biomass allometry

\subsubsection*{Function}

photosynthesis

\subsubsection*{Performance}

Growth rate: leaf expansion rate AND stem elongation rate
Biomass (show correlation between growth rate and biomass)

\cdm{Not sure where this goes} We tested for population variation in germination rate, measured as Days to Germination, using a lognormal survival model fit using the survreg function in the R package survival version 2.37-7 \citep{Therneau_2014}. The model was fit with Population as a fixed effect and Family as random effect using a $\Gamma$ frailty function. The signifcance of the Population effect was determined using analysis of deviance.

\subsection*{Intrinsic variation}

\cdm{which traits: growth rate, biomass, biomass allometry, leaf traits, photosynthesis?, mortality?}
We tested for Population, Treatment, and Population $\times$ Treatment interactions using mixed model ANOVA with Family included as a random factor. Models were fit by restricted maximum likelihood using lmer from the R package lme4 version 1.0-5 \citep{Bates_etal_2013}. Significant fixed effect terms were selected using a step-wise backward elimination procedure implemented with the step function in the R package lmerTest version 2.0-11 \citep{Kuznetsova_etal_2014}. Denominator degrees of freedom for $F$-tests were estimated using Satterthwaite's approximation. Significant Population effect indicate intrinsic trait differences; significant Population $\times$ Treatment effects indicate population differences in plasticity. For growth rate, we also accounted for differences in germination rate by including day of germination as a factor.

	%

\subsection*{Environmental correlates}
	% latitude
	% temperature (mean annual, July/june)
	% stream characterisitics - what data does Amy have from surveys?

	% approaches: direct correlation, bedazzle?

>>>>

\documentclass[11pt, oneside]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}

@@

% Package command for citing R packages
\newcommand{\pkg}[1]{{\fontseries{b}\selectfont #1}}

@@

%
% Where to find files
%

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
  \hlcom{# The path to the ms directory may need to be changed for different users}
  \hlcom{# Manuscript files}
  \hlstd{pathMS} \hlkwb{<-} \hlstr{"~/Google Drive/CardLocalAdaptation/ms"}

  \hlcom{# Data files for paper}
  \hlstd{pathDat} \hlkwb{<-} \hlkwd{paste}\hlstd{(pathMS,} \hlstr{"/Data"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}

  \hlcom{# Figures for paper}
  \hlstd{pathFig} \hlkwb{<-} \hlkwd{paste}\hlstd{(pathMS,} \hlstr{"/Figures"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}

  \hlcom{# Tables for paper}
  \hlstd{pathTab} \hlkwb{<-} \hlkwd{paste}\hlstd{(pathMS,} \hlstr{"/Tables"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}

  \hlcom{# Small function to make statistical significance asterisks}
  \hlstd{sigStar} \hlkwb{<-} \hlkwa{function}\hlstd{(}\hlkwc{pvalue}\hlstd{)}
  \hlstd{\{}
    \hlkwa{if} \hlstd{(pvalue} \hlopt{>=} \hlnum{0.05}\hlstd{)} \hlkwd{return}\hlstd{(}\hlstr{""}\hlstd{)}
    \hlkwa{if} \hlstd{(pvalue} \hlopt{<} \hlnum{0.05} \hlopt{&} \hlstd{pvalue} \hlopt{>=} \hlnum{0.01}\hlstd{)} \hlkwd{return}\hlstd{(}\hlstr{"*"}\hlstd{)}
    \hlkwa{if} \hlstd{(pvalue} \hlopt{<} \hlnum{0.01} \hlopt{&} \hlstd{pvalue} \hlopt{>=} \hlnum{0.001}\hlstd{)} \hlkwd{return}\hlstd{(}\hlstr{"**"}\hlstd{)}
    \hlkwa{if} \hlstd{(pvalue} \hlopt{<} \hlnum{0.001}\hlstd{)} \hlkwd{return}\hlstd{(}\hlstr{"***"}\hlstd{)}
  \hlstd{\}}

  \hlcom{# Color palette}
  \hlcom{# Primary color:}

   \hlstd{primShade0} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{1}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0.286}\hlstd{)}
   \hlstd{primShade1} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{1}\hlstd{,} \hlnum{0.643}\hlstd{,} \hlnum{0.745}\hlstd{)}
   \hlstd{primShade2} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{1}\hlstd{,} \hlnum{0.439}\hlstd{,} \hlnum{0.6}\hlstd{)}
   \hlstd{primShade3} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.765}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0.22}\hlstd{)}
   \hlstd{primShade4} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.588}\hlstd{,} \hlnum{0}\hlstd{,} \hlnum{0.169}\hlstd{)}

  \hlcom{# Secondary color (1):}

  \hlstd{sec1shade0} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.008}\hlstd{,} \hlnum{0.957}\hlstd{,} \hlnum{1}\hlstd{)}
  \hlstd{sec1shade1} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.643}\hlstd{,} \hlnum{0.984}\hlstd{,} \hlnum{1}\hlstd{)}
  \hlstd{sec1shade2} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.443}\hlstd{,} \hlnum{0.976}\hlstd{,} \hlnum{1}\hlstd{)}
  \hlstd{sec1shade3} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.49}\hlstd{,} \hlnum{0.514}\hlstd{)}
  \hlstd{sec1shade4} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0}\hlstd{,} \hlnum{0.376}\hlstd{,} \hlnum{0.396}\hlstd{)}

  \hlcom{# Secondary color (2):}

  \hlstd{sec2shade0} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.988}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{0}\hlstd{)}
  \hlstd{sec2shade1} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.996}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{0.643}\hlstd{)}
  \hlstd{sec2shade2} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.992}\hlstd{,} \hlnum{1}\hlstd{,} \hlnum{0.439}\hlstd{)}
  \hlstd{sec2shade3} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.835}\hlstd{,} \hlnum{0.847}\hlstd{,} \hlnum{0}\hlstd{)}
  \hlstd{sec2shade4} \hlkwb{<-} \hlkwd{rgb}\hlstd{(}\hlnum{0.643}\hlstd{,} \hlnum{0.651}\hlstd{,} \hlnum{0}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

@@

%
% Start document
%
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\maketitle
\listoffigures
\listoftables

@@

% write/analyze
% proposal for summer
  % local fitness set concave
  % range-wide fitness set convex
  % interaction between shape of fitness set and gene flow?
% evo-phys proposal

@@

Local adaptation is one of the most ubiquitous observations in nature: organisms perform well in their natal environment, but poorly outside it. Local adaptation within species most often involves genetic differences in ecologically relevant traits caused by selection to different environments, leading to predictable trait-environment correlations that persist in a common garden. Although local adaptation is well-supported by data from reciprocal transplant experiments, trait-environment associations, and ecological genetics/genomics, predicting which traits respond to selection exceeds our understanding. In particular, the role of physiological traits is poorly understood, even though physiology explains adaptation to the abiotic environment, which we need to understand as a baseline for anticipating how organisms will respond to climate change. A related problem is identifying the exact selective agents, such as abiotic factors like soil water availability or biotic interactions [etc.]... Here, we simultaneously investigate which traits and causal selective agents explain range-wide patterns of local physiological adaptation to climate in \textit{Mimulus}, a model genus for local adaptation studies.

% niche-tracking versus niche-shifting
One surprising pattern is that populations adapt to different environments by adjusting their life history to stay within the same fundamental physiological niche rather than have the niche itself evolve. For example, mosquitoes adjust diapause length with latitude rather than evolve altered ...\cdm{what other examples could we include here?} This suggests that in many situations it is evolutionarily easier to change life history rather than fundamental physiological tolerance.

% Selective agents
We also know little about the selective agents, such as climatic factors or biotic interactions, that are most important in driving adaptive divergence. In certain situations, such as pollinator shifts and edaphic adaptation, it is straightforward to identify putative selective agents and test predicted mechanisms. However, it is more challenging for cases in which traits vary with less specific factors like mean annual precipitation or latitude. For example, body size clines (larger size at higher latitude) have evolved repeatedly on three contients in \textit{Drosophila subobscura}, strongly indicating selection. But which one or more of the many climatic factors that covary with latitude actually controls selection on body size? Selection might be even more complex for species that live over multiple seasons, meaning that environmental variation rather than the average environment may be a selective agent.

% Motivate using Mimulus
Linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, genomic studies under both lab and field conditions. Many classic and contemporary studies of local adaptation have been conducted using species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources. ... Conspicuous absence of phys mechanism

% Final paragraph
1. `intrinsic' physiological variation
2. variation in stress response
3. what ecological factors explain divergence (i.e. putative selective agents underlying local adaptation)

Conclude that observations are best explained by stabilizing selection on the fundamental niche, but spatially varying divergent selection on coordinated life history and physiological traits.

@@

We used 16 populations from throughout the range of \textit{M. cardinalis} (Table 1). Seeds were collected in the field \cdm{Amy, can you explain seed collection methods?}.

\begin{table}[ht]
   \centering
   \topcaption{16 Focal populations}
   \begin{tabular}{@{} lllllll @{}}
      \toprule
  Name& Region  & Demo?& Pop gen?& Latitude  & Longtiude  & Elevation (mas) \\
      \midrule
	HAU & South Margin		& yes	& yes	& 	& 
    	&    \\
	CTC	& South Margin		& yes	& no	&  & 
    	&    \\
	CUR	& South Margin		& yes	& yes &  & 
    	&    \\
	GRP & South Margin		& no	& no	&  &
    	&    \\
	WWC &	Transverse		  & yes	& yes	&  & 
    	&    \\
	MIL	& Transverse	  	& yes	& no	&  & 
    	&    \\
	WFM	& Transverse	  	& yes	& no	&  & 
    	&    \\
	NMT	& South Sierras		& yes	& yes?&  & 
    	&    \\
	PRD	& South Sierras		& yes	& yes	&  & 
    	&    \\
	RWD	& South Sierras		& yes	& yes	&  & 
    	&    \\
	WNA	& Central Sierras	& yes	& yes	&  & 
    	&    \\
	RBW	& Central Sierras	& yes	& no	&  & 
    	&    \\
	MYU	& North Sierras		& yes	& yes	&  & 
    	&    \\
	LIJ	& North Sierras		& yes	& yes	&  & 
    	&    \\
	DPC	& North Coast	  	& yes	& yes	&  & 
    	&    \\
	RCC	& North Margin		& yes	& yes	&  & 
    	&    \\
	\bottomrule
   \end{tabular}
\end{table}

@@

On 14 April, 2014, 3-5 seeds per family were sown directly on sand (Quikrete Play Sand, Georgia, USA) watered to field capacity in RLC4 Ray Leach cone-tainers placed in RL98 98-well trays (Stuewe \& Sons, Inc., Oregon, USA). We used pure sand both to facilitate root-washing and because \textit{M. cardinalis} typically grows in sandy, riparian soils (A. Angert, pers. obs.). Two jumbo-sized cotton balls at the bottom of cone-tainers prevented sand from washing out. Cone-tainers were continuously bottom-watered during germination by placing them in medium-sized flow trays (FLOWTMD, Stuewe \& Sons, Inc., Oregon, USA) filled part way with water, placed on benches in greenhouses at the University British Columbia campus in Vancouver, Canada (49\degree 15' N, 123\degree 15' W). Misters thoroughly wetted the top of the sand every two hours during the day. Most seeds germinated between 1 and 2 weeks, but we allowed 3 weeks before transferring seedlings to growth chambers. Germination was recorded daily from one to two weeks after sowing, and every few days thereafter. On 5 May (21 days after sowing), seedlings were transferred to one of two MODEL Growth Chambers (Conviron, Manitoba, Canada). We thinned seedlings to one plant per cone-tainer, leaving the center-most, largest plant. 702 of 768 (91.4\%) had plants that could be used in the experiment. We allowed one week at constant, non stressful conditions (day: 20\celsius, night: 16\celsius) for plants to acclimate to growth chambers before starting treatments. The initial size of seedlings, measured as the length of the first true leaves, did not differ between populations, families, or treatments (Table S\#).

\paragraph*{Monitoring environment}

@@

We used 2 watering treatments on two populations on two soils. The watering treatments are: well-watered, drought.

@@

We simulated typical growing season (June and July) air temperatures at the two most thermally divergent focal sites in our study, Whitewater Canyon (High Temp) and Little Jameson (Low Temp). We downloaded daily interpolated mean, minimum, and maximum air temperature from 13 years (2000-2012) at both sites from ClimateWNA \citep{Wang_etal_2012}. Daily temperatures from ClimateWNA are usually highly correlated with the air temperature recorded from data loggers in the field at these sites (A. Angert, unpub. data). Hence, the ClimateWNA temperature profiles are likely to similar to actual thermal regimes experienced by \textit{M. cardinalis} in nature. To create realistic temperature regimes, we calculated the mean temperature trend from June to July using LOESS \citep{Cleveland_etal_1992}. The residuals were highly autocorrelated at both sites (warmer than average days are typically followed by more warm days) and there was strong correlation ($r = 0.65$) between sites (warm days in WWC were also warm in LIJ). The `VARselect' function in the \pkg{vars} package for R \citep{Pfaff_2008} indicated that a lag two Vector Autoregression (VAR(2)) model best captured the within-site autocorrelation as well as between-site correlation in residuals. We fit and simulated from the VAR(2) using the package \pkg{dse} \citep{Gilbert_2014} in R. Simulated data closely resembled the autocorrelation and between-site correlation of the actual data. From simulated mean temperature, we next selected minimum and maximum daily temperatures. Mean, min, and max temperature were highly correlated at both sites. We chose min and max temperatures using site-specific fitted linear models between mean, max, and min temperature, with additional variation given by normally-distributed random deviates with variance equal to the residual variance of the linear models. For each day, the nighttime (22:00 - 6:00) chamber temperature was set to the simulated minimum temperature. During the middle of the day, chamber temperature was set to the simulated maximum temperature, with a variable period of transition between min and max so that the average temperature was equal the simulated mean temperature.

@@

			\cdm{Are there any data from California about soil water content through the season?}

\subsubsection*{Growth and photosynthesis}

\begin{table}[ht]
   \centering
   \topcaption{Traits}
   \begin{tabular}{@{} ll @{}}
      \toprule
  Trait & Units \\
      \midrule
  Day of germination  & day \\
  Leaf expansion rate  &  mm day$^{-1}$  \\
  Shoot elongation rate  &  mm or cm day$^{-1}$  \\
  Harvest dry mass &  g  \\
  Photosynthetic rate &  $\mu$mol CO$_2$ m$^{-2}$ s$^{-1}$\\
  Mortality &  Probability?  \\
	    \bottomrule
   \end{tabular}
\end{table}

\paragraph{Day of germination} We tested for population variation in germination rate, measured as Days to Germination, using a lognormal survival model fit using the survreg function in the R package \pkg{survival} version 2.38 \citep{Therneau_2015}. The model was fit with Population as a fixed effect and Family as random effect using a $\Gamma$ frailty function. The signifcance of the Population effect was determined using analysis of deviance.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
  \hlkwd{library}\hlstd{(survival)}
  \hlstd{pathMS} \hlkwb{<-} \hlstr{"~/Google Drive/CardLocalAdaptation/ms"}
  \hlstd{pathDat} \hlkwb{<-} \hlkwd{paste}\hlstd{(pathMS,} \hlstr{"/Data"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}

  \hlstd{master} \hlkwb{<-} \hlkwd{read.csv}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/MasterDatasheet_out.csv"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{),}
    \hlkwc{row.names} \hlstd{=} \hlnum{1}\hlstd{)}

  \hlcom{# Create Surv object}
        \hlstd{germ} \hlkwb{<-} \hlkwd{with}\hlstd{(}\hlkwd{subset}\hlstd{(master,} \hlopt{!}\hlkwd{is.na}\hlstd{(master}\hlopt{$}\hlstd{MinGermDay)),} \hlkwd{Surv}\hlstd{(MinGermDay, MaxGermDay,}
                \hlkwc{type} \hlstd{=} \hlstr{"interval2"}\hlstd{))}

        \hlcom{# lognormal produced best fit to data}
        \hlstd{fitGerm} \hlkwb{<-} \hlkwd{survreg}\hlstd{(germ} \hlopt{~} \hlstd{PopID} \hlopt{+} \hlkwd{frailty}\hlstd{(Family,} \hlkwc{sparse} \hlstd{= F),}
                \hlkwc{data} \hlstd{=} \hlkwd{subset}\hlstd{(master,} \hlopt{!}\hlkwd{is.na}\hlstd{(master}\hlopt{$}\hlstd{MinGermDay)),} \hlkwc{dist} \hlstd{=} \hlstr{"lognormal"}\hlstd{)}
        \hlkwd{suppressWarnings}\hlstd{(}\hlkwd{anova}\hlstd{(fitGerm))} \hlcom{# significant family and population effects}
\end{alltt}
\begin{verbatim}
##                                   Df Deviance Resid. Df    -2*LL
## NULL                              NA       NA  700.0000 2796.417
## PopID                        15.0000 151.9495  685.0000 2644.468
## frailty(Family, sparse = F) 106.9738 254.7233  578.0262 2389.745
##                                 Pr(>Chi)
## NULL                                  NA
## PopID                       9.887489e-25
## frailty(Family, sparse = F) 4.530519e-14
\end{verbatim}
\end{kframe}
\end{knitrout}

\paragraph{Growth rate: leaf expansion and shoot elongation}

We censused longest leaf length $>1$ mm on 10 days (twice per week) between 12 May and 12 June (28 to 59 days after sowing). We ceased measuring leaf length once it appeared to asymptote and growth shifted to shoot elongation.  We also censused plant height on 7 days (twice per week) between 29 May and 20 June (45 to 67 days after sowing). Both leaf expansion and shoot elongation were modeled as a second-order polynomials of time with individual coefficients (separate for leaf and shoot growth) using empirical Bayes' estimates from linear mixed-effects models fit using the R package \pkg{lme4} version 1.1-7 \citep{Bates_etal_2014}.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
  \hlkwd{library}\hlstd{(lme4)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Loading required package: Matrix\\\#\# Loading required package: Rcpp}}\begin{alltt}
  \hlkwd{library}\hlstd{(lmerTest)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# \\\#\# Attaching package: 'lmerTest'\\\#\# \\\#\# The following object is masked from 'package:lme4':\\\#\# \\\#\#\ \ \ \  lmer\\\#\# \\\#\# The following object is masked from 'package:stats':\\\#\# \\\#\#\ \ \ \  step}}\begin{alltt}
  \hlcom{# New (blank) datasheet on which to add data}
  \hlstd{newdata} \hlkwb{<-} \hlkwd{read.csv}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/MasterDatasheet_in.csv"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}

  \hlcom{# Read in LER data}
  \hlstd{LERdata} \hlkwb{<-} \hlkwd{read.csv}\hlstd{(}\hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/LERdata.csv"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}

  \hlcom{#}
        \hlcom{# Empirical Bayes Approach: get coefficients for every plant}
        \hlcom{#}

        \hlcom{# suppressWarnings(mmLER <- lmer(log(LLL) ~ poly(DayN, 2, raw = T) + }
    \hlcom{# (poly(DayN, 2, raw = T)|indiv), data = LERdata))}
  \hlcom{# save(mmLER, file = paste(pathDat, "/mmLER.RData", sep = ""))}
  \hlkwd{load}\hlstd{(}\hlkwc{file} \hlstd{=} \hlkwd{paste}\hlstd{(pathDat,} \hlstr{"/mmLER.RData"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{))}

  \hlcom{# Extract coefficients and add to datasheet}
        \hlstd{Y} \hlkwb{<-} \hlkwd{t}\hlstd{(}\hlkwd{unlist}\hlstd{(}\hlkwd{fixef}\hlstd{(mmLER))} \hlopt{+} \hlkwd{t}\hlstd{(}\hlkwd{ranef}\hlstd{(mmLER)}\hlopt{$}\hlstd{indiv))[}\hlkwd{match}\hlstd{(newdata}\hlopt{$}\hlstd{indiv,}
                \hlkwd{rownames}\hlstd{(}\hlkwd{ranef}\hlstd{(mmLER)}\hlopt{$}\hlstd{indiv)), ]}
        \hlkwd{colnames}\hlstd{(Y)} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"b0_LER"}\hlstd{,} \hlstr{"b1_LER"}\hlstd{,} \hlstr{"b2_LER"}\hlstd{)}
  \hlstd{newdata} \hlkwb{<-} \hlkwd{cbind}\hlstd{(newdata, Y)}
